---
title: "Group Project 1"
subtitle: "Biology 368/664 Bucknell University"
output: html_notebook
author: Prof. Ken Field
date: 9 Feb 2022
---
#### Project Outline #####
Data set we are using: https://data.cdc.gov/NCHS/Provisional-COVID-19-Deaths-by-Sex-and-Age/9bhg-hcku/data

Outline:
1) Introduction to class
2) Introduction to R (M)
    Define RMD vs Regular R
    Loading Data Set
    Load Packages
    Refer to glossary for a function/stats overview…
3) Explore Data Set (with function explanations) (A)
    Formulate a question
    Read data
    Check packaging 
    Run structure
    Look at top and bottom of data
    Check your n’s
4) Visualize Data Set
    Timeline vs death (M)
    Rank net deaths by state (B - Coding Done)
    West coast vs east coast (A)
    Male female histogram with death (A)
    Do older people have higher death rates than younger (B)
5) Statistical analysis (linear model?) to answer hypothesis
    Age groups and death 
    Linear Model (B)
6) Glossary of Functions and statistical analyses (A+B+M)

#### Project Content #####



# Code for Ranking net Deaths by State

```{r}
library(readr)
Covid_19<-read_csv("Data/Provisional_COVID-19_Deaths_by_Sex_and_Age.csv")
#^Can be deleted once we cement our dataset

library(dplyr)

#Looking at original dataset and ensuring State = coded as a factor
Covid_19$State<-as_factor(Covid_19$State)
nlevels(Covid_19$State)
summary(Covid_19$State)
#54 states, need to remove non-states

#Filter functions (one to be used as an example of != and one to be used as an example of ==)
Covid_19.S<-filter(Covid_19, State != "United States" & State != "District of Columbia" & State != "New York City" & State != "Puerto Rico")
Covid_19.S.T.X.A<-filter(Covid_19.S, Group == "By Total" & Sex == "All Sexes" & `Age Group` == "All Ages")

#ensuring filter process worked
table(Covid_19.S.T.X.A$State)
nlevels(Covid_19.S.T.X.A$State)

#data display
arrange(Covid_19.S.T.X.A,desc(`COVID-19 Deaths`))
```






#### Project Details #####


This project will require you to develop a tutorial to teach Bucknell students how to use R for graphing and data analysis. 

## Target Audience

Discuss with your group the target audience for the tutorial. 
Examples could be one of the new core Biology classes, another 300-level course (not 364), or a research group. 

```{r echo=FALSE}

```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Load Libraries, include=FALSE}
# Load other packages here.
if (!require("datasauRus")) install.packages("datasauRus"); library(datasauRus)
if (!require("tidyverse")) install.packages("tidyverse"); library(tidyverse)
```

```{r}
library(readr)
Covid_19<-read_csv("./Data/Provisional_COVID-19_Deaths_by_Sex_and_Age.csv")
str(Covid_19)
```


## Grading

Each student will be expected to complete the following tasks to earn 85% of the points available for this assignment (21/25).

- Identify and obtain suitable dataset
- Use a Github repository and version control to collaborate on the project
- Spend 4-6 hours preparing, coding, and testing tutorial
  + Data exploration
  + Data visualization
  + Hypothesis testing
- Present tutorial in class
- Provide public archive suitable for sharing to students/faculty

Tutorials from previous classes can be viewed at our public github site: https://github.com/Bucknell-Biol364

Each group should use an *Acknowledgements* section to document the participation of each member and the collaboration within and between groups.

Additional credit will be awarded for providing assistance to other groups and for the development of a tutorial that goes beyond the minimal expectations listed above.



```{r}

```


```{r}

```

```{r}

```


```{r}

```



```{r}

```

# Acknowledgements


